A Constraint Solving Approach to Parikh Images of Regular Languages

### 2022-07-08 main.tex
how to find solutions

Parikh images play a central role in many automata-based algorithms,
for instance and notably in today's string solvers, which often have
to process constraints that combine regular language membership with
word length. To decide whether a simple formula like
$x \in \Language_1 \wedge y \in \Language_2 \wedge |x| > |y|$,
with string variables~$x, y$ and regular
languages~$\Language_1, \Language_2$, is satisfiable, it is
necessary to reason about the sets of word lengths induced by
$\Language_1, \Language_2$, which is a special case of the Parikh
image.  This required combined reasoning about strings and string
length has long been identified as a major bottleneck in string
solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,DBLP:journals/corr/BerzishZG17}.
Other string solvers make use of Parikh automata~\cite{parikh-automata}, and thus
Parikh images in the general case, to handle operations including
\verb!str.substr!  and \verb!str.at!, which comes at an even higher
price in terms of computational complexity~\cite{ostrich-plus}.

@@

It is possible to compute an existential Presburger formula describing
the Parikh image of any context-free language in linear
time~\cite{generate-parikh-image}; for the special case of regular
languages, this result was also stated in
\cite{muscholl-linear}. While theoretically elegant, this construction
has several disadvantages, often making it unpractical for integration
into algorithms. Firstly, the constructed Presburger formula contains
a linear number of existential quantifiers in the size of the
considered grammar, as well as complex Boolean structure, which is
needed to express the connectedness of sets of productions considered
in the construction. Eliminating those quantifiers to obtain a
quantifier-free representation of the Parikh image has exponential
complexity~\cite{DBLP:conf/issac/Weispfenning97}, and is in practice often impossible in reasonable
time. Just solving the Parikh image membership problem is NP-complete,
as it corresponds to computing a satisfying assignment of the
existential Presburger formula, and taxing for solvers as
well~\cite{ostrich-plus}.

@@

We assume basic familiarity with theorem provers, in particular sequent calculus
or similar. For reference, see e.g. \cite{Fitting96a}.

@@

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. Note that we operate on sets of
formulas. Additionally, we also use the convention of splitting the
formulas into linear equations and inequalities ($\SomeInequalities$)
and other formulas ($\SomeClause$). We assume that our predicates $\Image$ and $\Connected$ only occur positively.

The filtering function~$\Filter$ is evaluated symbolically, and can in practice
be read as a function from transitions to $\Naturals$-valued terms
(e.g.~\texttt{t} or~\texttt{t+1}). In our implementation \Catra{}, described in
\cref{sec:implementation}, $\Filter$~is a vector of fresh variables with the
same size as~$\Transitions$.

@@

Each of the rules of \cref{tbl:rules:single} should be read and
applied bottom-up, and relates
premises~$\SomeClause_1, \ldots, \SomeClause_k$ with some
conclusion~$\SomeClause$. When constructing a proof, we start from some
root~$\SomeClause$, and then apply proof rules to the goals of the proof
until a goal can be closed, or no more rule is applicable. Since
our rules operate by adding and matching linear (in)equalities in a proof goal,
we use the shorthand of listing the matched inequalities as antecedents. An
example of this can be seen in the~\Propagate{} rule. \Fudge{We pull solutions
out of a hat? I have no idea.}

>>>>

Parikh images play a central role in many automata-based algorithms,
for instance and notably in today's string solvers, which often have
to process constraints that combine regular language membership with
word length. To decide whether a simple formula like
$x \in \Language_1 \wedge y \in \Language_2 \wedge |x| > |y|$,
with string variables~$x, y$ and regular
languages~$\Language_1, \Language_2$, is satisfiable, it is
necessary to reason about the sets of word lengths induced by
$\Language_1, \Language_2$, which are special cases of the Parikh
image.  This required combined reasoning about strings and string
length has long been identified as a major bottleneck in string
solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,DBLP:journals/corr/BerzishZG17}.
Other string solvers make use of Parikh automata~\cite{parikh-automata}, and thus
Parikh images in the general case, to handle operations including
\verb!str.substr!  and \verb!str.at!, which comes at an even higher
price in terms of computational complexity~\cite{ostrich-plus}.

@@

It is possible to compute an existential Presburger formula describing
the Parikh image of any context-free language in linear
time~\cite{generate-parikh-image}; for the special case of regular
languages, this result was also stated in
\cite{muscholl-linear}. While theoretically elegant, this construction
has several disadvantages, often making it unpractical for integration
into algorithms. Firstly, the constructed Presburger formula contains
a linear number of existential quantifiers in the size of the
considered grammar, as well as complex Boolean structure, which is
needed to express the connectedness of sets of productions considered
in the construction. Eliminating those quantifiers to obtain a
quantifier-free representation of the Parikh image has exponential
complexity~\cite{DBLP:conf/issac/Weispfenning97}, and is often impossible in reasonable
time. Just solving the Parikh image membership problem is NP-complete,
as it corresponds to computing a satisfying assignment of the
existential Presburger formula, and taxing for solvers as
well~\cite{ostrich-plus}.

@@

We assume basic familiarity with first-order logic, Presburger
arithmetic, and the classical sequent calculus. For reference, see
e.g. \cite{Fitting96a}.

@@

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. The rules operate on sets of formulas, and
can be interpreted as rules of a one-sided sequent calculus, in
which all formulas are located in the antecedent~\cite{Fitting96a}.
Each of the rules should be read and
applied bottom-up, and relates
premises~$\SomeClause_1, \ldots, \SomeClause_k$ with some
conclusion~$\SomeClause$. When constructing a proof, we start from some
root~$\SomeClause$, and then apply proof rules to the goals of the proof
until a goal can be closed, or no more rule is applicable.

A proof in which all proof goals could be closed shows that the formulas
in the root~$\SomeClause$ of the proof are inconsistent, and do not
have any solutions. A proof goal that could not be closed, but to which no
more rules are applicable, gives rise to  a solution of the formulas in the
root~$\SomeClause$. Such an open proof goal will only contain formulas
in Presburger arithmetic, so that a solution can be computed using
standard algorithms~\cite{Fitting96a}.

We use the convention of splitting the
formulas into linear (in)equalities ($\SomeInequalities$)
and other formulas ($\SomeClause$), and assume that predicates
$\Image$ and $\Connected$ only occur positively.
 Since
our rules operate by adding and matching linear (in)equalities in a proof goal,
we use the shorthand of listing the matched inequalities as antecedents
e.g., in \Propagate{}).
The filtering function~$\Filter$ is evaluated symbolically, and can in practice
be read as a function from transitions to $\Naturals$-valued terms
(e.g.~\texttt{t} or~\texttt{t+1}). In our implementation \Catra{}, described in
\cref{sec:implementation}, $\Filter$~is a vector of fresh variables with the
same size as~$\Transitions$.

@@
